Reference Tracking Strategies of Deaf Adult Signers in Turkish Sign Language

### 2023-01-31 revised_manuscript.Rmd
update

However, this particular observation is not peculiar to spoken languages. There is an emerging body of literature that reports the distribution of REs in signed narratives [@czubek2017; @swabey2002; @frederiksen2016; @perniss2015; @morgan2005; @hodge2019; @ahlgren1994]. Building on the work by @gundel1993, @swabey2002 demonstrated that, like other languages such as Russian, Chinese, and Japanese, American Sign Language (ASL) signers use bare nouns for introduced and re-introduced discourse contexts whereas zero anaphora is used for referents which are "in-focus" or very accessible to the addressee. Overt REs like pronouns or nouns are disprefered for such referents. @swabey2002 concluded that "native" signers in her study abided by the Gricean maxim of quantity and thus used more economic forms of reference tracking. Similar findings have been reported for German Sign Language (Deutsche Gebärdensprache - DGS) [@perniss2015] and British Sign Language (BSL) [@morgan2005]. It is important to note, however, that this research either collapsed DoD and DoH signers into a single group and labeled them as "native-like" or explicitly compared native deaf signers with late deaf signers (with delayed exposure to a sign language) in terms of narrative competence.

@@

```{=tex}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:5a} Constructed action \\\glll
{} \includegraphics[width=40pt]{pictures/p10.png} \\ \textbf{Gloss:} \textsc{mouse.stay.like.this} \\ 
\textbf{Status:} Maintained. \\
\glt \textbf{Context:} Narration 6, Event 9: The signer re-introduces Jerry in the previous clause and maintains reference to him by using this constructed action.
\glt \textbf{Translation:} (The mouse) stays like this.
\vspace{0.35cm}

\ex \label{ex:5b} Plain verb \\\glll
{} \includegraphics[width=80pt]{pictures/p11.png} \\ \textbf{Gloss:} \textsc{run.away} \\ 
\textbf{Status:} Re-introduced. \\
\glt \textbf{Context:} Narration 5, Event 4: In this sentence, the signer re-introduces Jerry with the plain verb \textsc{run.away}.
\glt \textbf{Translation:} (Jerry) runs away.
\vspace{0.35cm}

\ex \label{ex:5c} Agreement verb \\\glll
{} \includegraphics[width=80pt]{pictures/p12.png} \\ \textbf{Gloss:} \textsc{3a(mouse)-hit-3b(cat)} \\ 
\textbf{Status:} Maintained. \\
\glt \textbf{Context:} Narration 8, Event 2: The signer introduces Jerry in the preceding clause and maintains reference to him by using the agreement verb \textsc{hit}.
\glt \textbf{Translation:} (Jerry) hits (Tom).
\vspace{0.35cm}

@@

For age of acquisition effects, based on what is proposed for child and adult deaf signers with delayed language exposure in terms of their morphosyntactic [@boudreault2006; @cormier2012; @kayaba2022] and narrative development [@becker2009; @cormier2013], we expected that DoH signers would be more over-redundant in their referent tracking, akin to second language learners [@frederiksen2019; @bel2015]. That is, we had the expectation that DoH signers would use more explicit markers like nouns and fewer implicit markers like zero anaphora even for highly accessible referents (when the discourse status is maintained or even re-introduced in same cases) than DoD signers.

@@

As for the distribution of REs across different discourses statuses, we observed that signers overwhelmingly preferred to use nominals over other REs for the first mention of a referent and used zero anaphora such as CA, agreement and plain verb for referents with high accessibility (i.e., when the discourse status was maintained). This finding parallels other accounts of referential accessibility suggested for other sign languages (see @frederiksen2016; @czubek2017; and @swabey2002; for ASL, @morgan2005; for BSL @bel2015; and @barbera2009 for LSC). Our results also indicate that modified nouns (e.g., nouns that are preceded or followed by a classifier or IX) appeared more in introduced contexts compared to re-introduced contexts. This indicates that modified nouns are used more to track referents with lower accessibility, probably due to their more explicit structure.

>>>>

However, this particular observation is not peculiar to spoken languages. There is an emerging body of literature that reports the distribution of REs in signed narratives [@czubek2017; @ferrara2022; @swabey2002; @frederiksen2016; @perniss2015; @morgan2005; @hodge2019; @ahlgren1994]. Building on the work by @gundel1993, @swabey2002 demonstrated that, like other languages such as Russian, Chinese, and Japanese, American Sign Language (ASL) signers use bare nouns for introduced and re-introduced discourse contexts whereas zero anaphora is used for referents which are "in-focus" or very accessible to the addressee. Overt REs like pronouns or nouns are disprefered for such referents. @swabey2002 concluded that "native" signers in her study abided by the Gricean maxim of quantity and thus used more economic forms of reference tracking. Similar findings have been reported for German Sign Language (Deutsche Gebärdensprache - DGS) [@perniss2015] and British Sign Language (BSL) [@morgan2005]. It is important to note, however, that this research either collapsed DoD and DoH signers into a single group and labeled them as "native-like" or explicitly compared native deaf signers with late deaf signers (with delayed exposure to a sign language) in terms of narrative competence. Most recently, @ferrara2022 examined corpora including video-recorded narratives from five different sign languages including Irish Sign Language, Finnish Sign Language, Norwegian Sign Language and Swedish Sign Language and investigated how these languages established and maintained referents. Like other observations, this research showed that, across each of these languages, bare nouns were used to introduce new referents and zero anaphora or more reduced forms were preferred for referent maintenance.

@@

```{=tex}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:5a} Constructed action \\\glll
{} \includegraphics[width=40pt]{pictures/p10.png} \\ \textbf{Gloss:} \textsc{mouse.stay.like.this} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 6, Event 9: The signer re-introduces Jerry in the previous clause and maintains reference to him by using this constructed action.
\glt \textbf{Translation:} (The mouse) stays like this.
\vspace{0.35cm}

\ex \label{ex:5b} Plain verb \\\glll
{} \includegraphics[width=80pt]{pictures/p11.png} \\ \textbf{Gloss:} \textsc{run.away} \\ 
\textbf{Status:} Re-introduce. \\
\glt \textbf{Context:} Narration 5, Event 4: In this sentence, the signer re-introduces Jerry with the plain verb \textsc{run.away}.
\glt \textbf{Translation:} (Jerry) runs away.
\vspace{0.35cm}

\ex \label{ex:5c} Agreement verb \\\glll
{} \includegraphics[width=80pt]{pictures/p12.png} \\ \textbf{Gloss:} \textsc{3a(mouse)-hit-3b(cat)} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 8, Event 2: The signer introduces Jerry in the preceding clause and maintains reference to him by using the agreement verb \textsc{hit}.
\glt \textbf{Translation:} (Jerry) hits (Tom).
\vspace{0.35cm}

@@

For age of acquisition effects, based on what is proposed for child and adult deaf signers with delayed language exposure in terms of their morphosyntactic [@boudreault2006; @cormier2012; @kayaba2022] and narrative development [@becker2009; @cormier2013], we expected that DoH signers would be more over-redundant in their referent tracking, perhaps akin to second language learners [@frederiksen2019; @bel2015]. That is, we had the expectation that DoH signers would use more explicit markers like nouns and fewer implicit markers like zero anaphora even for highly accessible referents (when the discourse status is maintained or even re-introduced in same cases) than DoD signers.

@@

As for the distribution of REs across different discourses statuses, we observed that signers overwhelmingly preferred to use nominals over other REs for the first mention of a referent and used zero anaphora such as CA, agreement and plain verb for referents with high accessibility (i.e., when the discourse status was maintained). This finding parallels other accounts of referential accessibility suggested for other sign languages (see @frederiksen2016; @czubek2017; and @swabey2002; for ASL, @morgan2005; for BSL @bel2015; @barbera2009 for LSC; @ferrara2022 for other sign languages). Our results also indicate that modified nouns (e.g., nouns that are preceded or followed by a classifier or IX) appeared more in introduced contexts compared to re-introduced contexts. This indicates that modified nouns are used more to track referents with lower accessibility, probably due to their more explicit structure.

### 2023-01-31 revised_manuscript.tex
update

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives (Ahlgren and Bergman, 1994;
Czubek, 2017; Frederiksen and Mayberry, 2016; Hodge et al., 2019;
Morgan, 2005; Perniss and Özyürek, 2015; Swabey, 2002). Building on the
work by Gundel et al. (1993), Swabey (2002) demonstrated that, like
other languages such as Russian, Chinese, and Japanese, American Sign
Language (ASL) signers use bare nouns for introduced and re-introduced
discourse contexts whereas zero anaphora is used for referents which are
``in-focus'' or very accessible to the addressee. Overt REs like
pronouns or nouns are disprefered for such referents. Swabey (2002)
concluded that ``native'' signers in her study abided by the Gricean
maxim of quantity and thus used more economic forms of reference
tracking. Similar findings have been reported for German Sign Language
(Deutsche Gebärdensprache - DGS) (Perniss and Özyürek, 2015) and British
Sign Language (BSL) (Morgan, 2005). It is important to note, however,
that this research either collapsed DoD and DoH signers into a single
group and labeled them as ``native-like'' or explicitly compared native
deaf signers with late deaf signers (with delayed exposure to a sign
language) in terms of narrative competence.

@@

\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:5a} Constructed action \\\glll
{} \includegraphics[width=40pt]{pictures/p10.png} \\ \textbf{Gloss:} \textsc{mouse.stay.like.this} \\ 
\textbf{Status:} Maintained. \\
\glt \textbf{Context:} Narration 6, Event 9: The signer re-introduces Jerry in the previous clause and maintains reference to him by using this constructed action.
\glt \textbf{Translation:} (The mouse) stays like this.
\vspace{0.35cm}

\ex \label{ex:5b} Plain verb \\\glll
{} \includegraphics[width=80pt]{pictures/p11.png} \\ \textbf{Gloss:} \textsc{run.away} \\ 
\textbf{Status:} Re-introduced. \\
\glt \textbf{Context:} Narration 5, Event 4: In this sentence, the signer re-introduces Jerry with the plain verb \textsc{run.away}.
\glt \textbf{Translation:} (Jerry) runs away.
\vspace{0.35cm}

\ex \label{ex:5c} Agreement verb \\\glll
{} \includegraphics[width=80pt]{pictures/p12.png} \\ \textbf{Gloss:} \textsc{3a(mouse)-hit-3b(cat)} \\ 
\textbf{Status:} Maintained. \\
\glt \textbf{Context:} Narration 8, Event 2: The signer introduces Jerry in the preceding clause and maintains reference to him by using the agreement verb \textsc{hit}.
\glt \textbf{Translation:} (Jerry) hits (Tom).
\vspace{0.35cm}

@@

For age of acquisition effects, based on what is proposed for child and
adult deaf signers with delayed language exposure in terms of their
morphosyntactic (Boudreault and Mayberry, 2006; Cormier et al., 2012;
Kayabaşı and Gökgöz, 2022) and narrative development (Becker, 2009;
Cormier et al., 2013), we expected that DoH signers would be more
over-redundant in their referent tracking, akin to second language
learners (Bel et al., 2015; Frederiksen and Mayberry, 2019). That is, we
had the expectation that DoH signers would use more explicit markers
like nouns and fewer implicit markers like zero anaphora even for highly
accessible referents (when the discourse status is maintained or even
re-introduced in same cases) than DoD signers.

@@

As for the distribution of REs across different discourses statuses, we
observed that signers overwhelmingly preferred to use nominals over
other REs for the first mention of a referent and used zero anaphora
such as CA, agreement and plain verb for referents with high
accessibility (i.e., when the discourse status was maintained). This
finding parallels other accounts of referential accessibility suggested
for other sign languages (see Frederiksen and Mayberry (2016); Czubek
(2017); and Swabey (2002); for ASL, Morgan (2005); for BSL Bel et al.
(2015); and Barberà and Massó (2009) for LSC). Our results also indicate
that modified nouns (e.g., nouns that are preceded or followed by a
classifier or IX) appeared more in introduced contexts compared to
re-introduced contexts. This indicates that modified nouns are used more
to track referents with lower accessibility, probably due to their more
explicit structure.

@@

>>>>

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives (Ahlgren and Bergman, 1994;
Czubek, 2017; Ferrara et al., 2022; Frederiksen and Mayberry, 2016;
Hodge et al., 2019; Morgan, 2005; Perniss and Özyürek, 2015; Swabey,
2002). Building on the work by Gundel et al. (1993), Swabey (2002)
demonstrated that, like other languages such as Russian, Chinese, and
Japanese, American Sign Language (ASL) signers use bare nouns for
introduced and re-introduced discourse contexts whereas zero anaphora is
used for referents which are ``in-focus'' or very accessible to the
addressee. Overt REs like pronouns or nouns are disprefered for such
referents. Swabey (2002) concluded that ``native'' signers in her study
abided by the Gricean maxim of quantity and thus used more economic
forms of reference tracking. Similar findings have been reported for
German Sign Language (Deutsche Gebärdensprache - DGS) (Perniss and
Özyürek, 2015) and British Sign Language (BSL) (Morgan, 2005). It is
important to note, however, that this research either collapsed DoD and
DoH signers into a single group and labeled them as ``native-like'' or
explicitly compared native deaf signers with late deaf signers (with
delayed exposure to a sign language) in terms of narrative competence.
Most recently, Ferrara et al. (2022) examined corpora including
video-recorded narratives from five different sign languages including
Irish Sign Language, Finnish Sign Language, Norwegian Sign Language and
Swedish Sign Language and investigated how these languages established
and maintained referents. Like other observations, this research showed
that, across each of these languages, bare nouns were used to introduce
new referents and zero anaphora or more reduced forms were preferred for
referent maintenance.

@@

\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:5a} Constructed action \\\glll
{} \includegraphics[width=40pt]{pictures/p10.png} \\ \textbf{Gloss:} \textsc{mouse.stay.like.this} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 6, Event 9: The signer re-introduces Jerry in the previous clause and maintains reference to him by using this constructed action.
\glt \textbf{Translation:} (The mouse) stays like this.
\vspace{0.35cm}

\ex \label{ex:5b} Plain verb \\\glll
{} \includegraphics[width=80pt]{pictures/p11.png} \\ \textbf{Gloss:} \textsc{run.away} \\ 
\textbf{Status:} Re-introduce. \\
\glt \textbf{Context:} Narration 5, Event 4: In this sentence, the signer re-introduces Jerry with the plain verb \textsc{run.away}.
\glt \textbf{Translation:} (Jerry) runs away.
\vspace{0.35cm}

\ex \label{ex:5c} Agreement verb \\\glll
{} \includegraphics[width=80pt]{pictures/p12.png} \\ \textbf{Gloss:} \textsc{3a(mouse)-hit-3b(cat)} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 8, Event 2: The signer introduces Jerry in the preceding clause and maintains reference to him by using the agreement verb \textsc{hit}.
\glt \textbf{Translation:} (Jerry) hits (Tom).
\vspace{0.35cm}

@@

For age of acquisition effects, based on what is proposed for child and
adult deaf signers with delayed language exposure in terms of their
morphosyntactic (Boudreault and Mayberry, 2006; Cormier et al., 2012;
Kayabaşı and Gökgöz, 2022) and narrative development (Becker, 2009;
Cormier et al., 2013), we expected that DoH signers would be more
over-redundant in their referent tracking, perhaps akin to second
language learners (Bel et al., 2015; Frederiksen and Mayberry, 2019).
That is, we had the expectation that DoH signers would use more explicit
markers like nouns and fewer implicit markers like zero anaphora even
for highly accessible referents (when the discourse status is maintained
or even re-introduced in same cases) than DoD signers.

@@

As for the distribution of REs across different discourses statuses, we
observed that signers overwhelmingly preferred to use nominals over
other REs for the first mention of a referent and used zero anaphora
such as CA, agreement and plain verb for referents with high
accessibility (i.e., when the discourse status was maintained). This
finding parallels other accounts of referential accessibility suggested
for other sign languages (see Frederiksen and Mayberry (2016); Czubek
(2017); and Swabey (2002); for ASL, Morgan (2005); for BSL Bel et al.
(2015); Barberà and Massó (2009) for LSC; Ferrara et al. (2022) for
other sign languages). Our results also indicate that modified nouns
(e.g., nouns that are preceded or followed by a classifier or IX)
appeared more in introduced contexts compared to re-introduced contexts.
This indicates that modified nouns are used more to track referents with
lower accessibility, probably due to their more explicit structure.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ferrara2022}{}}%
Ferrara, Lindsay, Anible, Benjamin, Hodge, Gabrielle, Jantunen, Tommi,
Leeson, Lorraine, Mesch, Johanna, Nilsson, Anna-Lena, 2022. A
cross-linguistic comparison of reference across five signed languages.
Linguistic Typology.
